---
title: "Accessing NCBI data with the Rentrez package"
subtitle: "R Community of Practice June 15 - July 20"
author: "Irmarie Fraticelli-Rodriguez"
date: "07/13/2023"

format:
  revealjs:
   theme: [default, custom.scss]
   slide-number: true
   smaller: true
   scrollable: true
   preview-links: true
   code-fold: show
   code-overflow: wrap
   auto-play-media: true
footer: <https://www2.hshsl.umaryland.edu/cdabs/>
---
```{r}
library(rentrez)
```


## Learning Goals:
**This section objectives are:**

::: {.r-fit-text}
1. Become familiar with `Rentrez` and its basic search functions.
1. Wrangle data to a usable form.
::: 

## The Data

::: {.r-fit-text}
In previous weeks, we have been working with data that exists locally, in your computers, and are in a `CSV` format. 

Not all data is formatted in a `CSV`. Sometimes we may need to connect directly with databases to extract data and we may get it in other formats like `JSON` or `XML`. 

Today our data comes from PubMed, a National Center for Biotechnology Information (NCBI) database. 

As a way to interact with this database and perform a search query, we are going to use `Rentrez`. 

**Output of this session:**

1. Download `Rentrez` and learn basic search functions.
1. Create a graph or document using your search results. 
:::

## What is Rentrez? 

`Rentrez` is an R interface that allows its users to interact with NCBI API.

With Rentrez, you do not need to use any additional program or terminal to access NCBI Data. 

This means that you can request data from multiple databases (PubMed, SNP, Clinvar, SRA, Gene, and others) in the same RStudio Session.


## Rentrez Functions 

::: {.r-fit-text}
The functions listed below help you learn a little bit more about `Rentrez` and NCBI Databases.

| Function                | Usage                                                |
|-------------------------|------------------------------------|
| entrez_dbs()            | List of Entrez Databases                             |
| entrez_db_summary()     | Brief description of what the database is            |
| entrez_db_searchable()  | Set of search fields that can used with this database |


:::

## Performing searches in NCBI API

::: {.r-fit-text}
The functions listed below help you perform searches in a NCBI Database.

| Function                | Usage                              |                 
|-------------------------|------------------------------------|
| entrez_search()         | Allows you to perform simple or boolean searches |
| entrez_summary()        | Allows you to retrieve basic information about the records found |
| extract_from_esummary() | Extract elements from a list of esummary records     |

:::

## entrez_search()

Similar to PubMed, `Rentrez` allows you to perform simple or boolean searches using the same structure that you would use in the PubMed search bar. The allowed boolean terms are **AND, OR, and NOT**.

<br/>

First, let's learn the syntax for a simple and a boolean search:

<br/>

**Simple Search:**

```{r}
#| echo: true

pcos_pm <- entrez_search(db="pubmed", term = "pcos[all]")

```


**Boolean search:**

```{r}
#| echo: true

pcos_ir_pm <- entrez_search(db="pubmed", term= "pcos[all] AND insulin resistance[all]")

```

## entrez_summary()

Now let's retrieve basic information about the records we collected in one of our searches. But, first, let's learn the syntax of the `entrez_summary()` function.

```{r}
#| echo: true

summary_pcos_pm <- entrez_summary(db="pubmed", id=pcos_pm$ids)

```

## extract_from_esummary()

This function helps you navigate through an XMLInternalDocument and extract elements from a list of esummary records. But, first, let's learn the syntax of the `extract_from_esummary()` function.

<br/>

```{r}
#| echo: true

uids <- extract_from_esummary(summary_pcos_pm,"uid")
authors <- extract_from_esummary(summary_pcos_pm,"authors")["name",]
pubdate <- extract_from_esummary(summary_pcos_pm,"pubdate")

```


## Exercise: Let's practice!

Now lets perform our own searches!

We are going to use the `Rentrez` functions to extract data from an NCBI database. 

Our objective is to: 

1. Build your query: Identify your search terms, adequate database, search fields and perform the search using rentrez.

2. Get article/object summaries.  

3. Select the values that you would like to save (for example author, title, source).

4. Select one of two outputs: 
  - Create a graph that represents the results of your search.
  - Move CSV or any other type of document appropriate for your search. 

## **STEP #1**

Build your query: Identify your search terms, adequate database, and search fields and perform the search using `Rentrez`.    

```{r}

pcos_pm <- entrez_search(db="pubmed", term = "pcos[all]")

```


## **STEP #2**  

Get article summaries

```{r}

summary_pcos_pm <- entrez_summary(db="pubmed", id=pcos_pm$ids)

```


## **STEP #3**

Select the values that you would like to save (for example author, title, source)

```{r}

uids <- extract_from_esummary(summary_pcos_pm,"uid")
authors <- extract_from_esummary(summary_pcos_pm,"authors")["name",]
pubdate <- extract_from_esummary(summary_pcos_pm,"pubdate")
title <- extract_from_esummary(summary_pcos_pm,"title")
pubtype <- extract_from_esummary(summary_pcos_pm,"pubtype")
source <- extract_from_esummary(summary_pcos_pm,"source")
issue <- extract_from_esummary(summary_pcos_pm,"issue")
volume <- extract_from_esummary(summary_pcos_pm,"volume")
pages <- extract_from_esummary(summary_pcos_pm,"pages")
articleids <- extract_from_esummary(summary_pcos_pm,"articleids", simplify=FALSE)
  
```


## **STEP #4** 
- Select one of two outputs: 
  - Create a graph that represents the results of your search.
  - Move CSV or any other type of document appropriate for your search. 
```{r}
# Creating a CSV or YAML
install.packages("yaml")
library(yaml)
library(tidyverse)

# Include Non Atomic Fields (author, pubtype)
yaml_data <- tibble(authors,title,pubtype,source,issue,volume,pages,pubdate,uids)

# Exclude Non Atomic Fields (author, pubtype)
csv_data <- tibble(title,source,issue,volume,pages,pubdate,uids)

# Store in YAML and CSV formats respectively
write_yaml(transpose(yaml_data),"pcos.yaml")
write_csv(csv_data,"pcos.csv")

```
