---
title: "R Community of Practice"
author: ""
format: 
  revealjs:
   code-fold: show
---

## What data do you chose to work with and why?

During the six week of the [R Community of Practice](https://hshsl-training.github.io/RCOP/), I chose to use *PubMed* data.
I decided to learn how to use the R package `Rentrez` to extract data from the PubMed database. Also, I learn to create data visualizations using this package.  

## What questions did you had about the data that you wanted to answer in the R community of practice? 

My questions for *PubMed* data were:

- How can I get PubMed data without leaving my RStudio window?
- Can I extract publication dates and journals from the PubMed data using Rentrez?
  - How can I display this data in a graph?

## What steps did you to take to process the data? 

To get data from the *PubMed* database using `Rentrez`, I had to perform the following steps: 

1. Read `Rentrez` documentation
1. Install and load `Rentrez` package
1. Perform my first search in *PubMed* using the command `entrez_search()` 
1. Get summaries of the articles found in my search using `entrez_summary`
1. Extract interesting and needed items using `extract_from_esummary()`

## A graphical summary of what they found out? 

With the skills learned at the R Community of Practice, I was able to create a graph using `PubMed` Data. 

```{r echo=TRUE}
library(rentrez)
library(ggplot2)

#Performing Search
search <- entrez_search(db = "pubmed", term = "Kristen A. Stafford[AUTH]", retmax=100)

#Getting all the summaries
summary <- entrez_summary(db = "pubmed",id=search$ids)

#Extracting information that I am interested in
extract <- extract_from_esummary(summary, "source")

```


---

```{r echo=TRUE}
library(rentrez)
library(ggplot2)

#Creating a data visualization
ggplot(mapping = aes(x=extract)) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "Kristen's Stafford Publications per journal",
    x="Journal", y= "Amount")
```


